R for Clinical Study Reports and Submission

### 2021-08-27 tlf-overview.Rmd
tlf-overview.Rmd edited online with Bitbucket
update tlf-overview page

>>>>

## Tools

In this part, we mainly use R packages below to illustrate 
how to deliver TLFs in CSR.

- [tidyverse](https://www.tidyverse.org/): prepare reporting ready datasets. 
- [r2rtf](https://merck.github.io/r2rtf/): create RTF outputs

### tidyverse

`tidyverse` is a collection of R packages to simplify the workflow to manipulate, 
visualize and analyze data in R. 
Those R packages share [the tidy tools manifesto](https://cran.r-project.org/web/packages/tidyverse/vignettes/manifesto.html) 
and easy to use for interactive data analysis.

RStudio provided outstanding [cheatsheets](https://www.rstudio.com/resources/cheatsheets/), 
and [tutorials](https://github.com/rstudio-education/remaster-the-tidyverse) for `tidyverse`.

There are also books to introduce tidyverse. 
We assume reader to have experience in using `tidyverse` in this book.

- [The tidyverse cookbook](https://rstudio-education.github.io/tidyverse-cookbook/)
- [R for Data Science](https://r4ds.had.co.nz/)

### r2rtf

`r2rtf` is an R package to create production ready tables and figures in RTF format.
The R package is designed to

- provide simple "verb" functions that correspond to each component of a table, 
  to help you translate data frame to table in RTF file.
- enable pipes (`%>%`). 
- only focus on **table format**. 
  Data manipulation and analysis shall be handled by other R packages. (e.g. `tidyverse`)

Before creating an RTF table we need to:

- Figure out table layout.

- Split the layout into small tasks in the form of a computer program.

- Execute the program.

We provided a brief introduction of `r2rtf` and show how to transfer 
data frames into Table, Listing, and Figure (TLFs).

Other extended examples and features are covered in 
[`r2rtf` package website](https://merck.github.io/r2rtf/articles/index.html).

To explore the basic RTF generation verbs in `r2rtf`, 
we'll use the dataset `r2rtf_adae` saved in the `r2rtf` package. 
This dataset contain adverse events (AE) information from a clinical trial.

Below is the meaning of relevant variables. 
More information can be found in help page of the dataset (`?r2rtf_adae`)

In this example we consider three variables:

- USUBJID: Unique Subject Identifier
- TRTA: Actual Treatment 
- AEDECOD: Dictionary-Derived Term

```{r}
library(r2rtf)
r2rtf_adae %>%
  select(USUBJID, TRTA, AEDECOD) %>%
  head(4)
```
`dplyr` and `tidyr` packages within `tidyverse` are used 
for data manipulation to create a data frame 
that contains all the information we want to add in an RTF table.

```{r}
library(tidyr)
library(dplyr)
tbl <- r2rtf_adae %>%
  count(TRTA, AEDECOD) %>%
  pivot_wider(names_from = TRTA, values_from = n, values_fill = 0)

tbl %>% head(4)
```

Now we have a dataset `tbl` in preparing the final RTF table.

`r2rtf` aims to provide one function for each type of table layout.
Commonly used verbs includes:

- `rtf_page`: RTF page information
- `rtf_title`: RTF title information
- `rtf_colheader`: RTF column header information
- `rtf_body`: RTF table body information
- `rtf_footnote`: RTF footnote information
- `rtf_source`: RTF data source information

All these verbs is designed to enables usage of pipes (`%>%`).
A full list of all functions can be found in [package reference](https://merck.github.io/r2rtf/reference/index.html)

A minimal example below illustrates how to combine verbs using pipes to create an RTF table.

- `rtf_body` is used to define table body layout. 
- `rtf_encode` transfers table layout information into RTF syntax.
- `write_rtf` save RTF encoding into a file with file extension `.rtf`

```{r}
head(tbl) %>%
  rtf_body() %>%                        # Step 1 Add table  attributes
  rtf_encode() %>%                      # Step 2 Convert attributes to RTF encode
  write_rtf("tlf/intro-ae1.rtf")        # Step 3 Write to a .rtf file
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/intro-ae1.pdf")
```

If we want to adjust the width of each column to 
provide more space to the first column, 
this can be achieved by updating `col_rel_width` argument in `rtf_body`.

In this example, the input of `col_rel_width` is a vector with same length for number of columns. 
This argument defines the relative width of each column 
within a pre-defined total column width.

In this example, the defined relative width is `3:2:2:2`. 
Only the ratio of `col_rel_width` is used. 
Therefore it is equivalent to use `col_rel_width = c(6,4,4,4)` 
or `col_rel_width = c(1.5,1,1,1)`.

```{r}
head(tbl) %>%
  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%    # define relative width
  rtf_encode() %>%
  write_rtf("tlf/intro-ae2.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/intro-ae2.pdf")
```

In previous example, we found issue of misaligned column header. 
We can fix the issue by using `rtf_colheader` function.

In `rtf_colheader`, `colheader` argument is used to provide content of column header. 
We use `"|"` to separate the columns.

In the example below, `"Adverse Events | Placebo | Xanomeline High Dose | Xanomeline Low Dose"`
define a column header with 4 columns.

```{r}
head(tbl) %>%
  rtf_colheader(
    colheader = "Adverse Events | Placebo | Xanomeline High Dose | Xanomeline Low Dose",
    col_rel_width = c(3, 2, 2, 2)
  ) %>%
  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%
  rtf_encode() %>%
  write_rtf("tlf/intro-ae3.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/intro-ae3.pdf")
```

In `rtf_*` functions such as `rtf_body`, `rtf_footnote`, 
`text_justification` argument is used to align text. 
Default is `"c"` for center justification. 
To vary text justification by column, use character vector with length of vector equal to
number of columns displayed (e.g., `c("c","l","r")`).

All possible inputs can be found in the table below.

```{r}
r2rtf:::justification()
```

Below is an example to make first column left aligned and center aligned for the rest.

```{r}
head(tbl) %>%
  rtf_body(text_justification = c("l", "c", "c", "c")) %>%
  rtf_encode() %>%
  write_rtf("tlf/intro-ae5.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/intro-ae5.pdf")
```

In `rtf_*` functions such as `rtf_body`, `rtf_footnote`, etc.,
`border_left`, `border_right`, `border_top`, and `border_bottom` control cell borders.

If we want to remove the top border of `"Adverse Events"` in header, 
we can change default value `"single"` to `""` in `border_top` argument as showed below.

`r2rtf` support 26 different border types. The details can be found in 
the [r2rtf package website](https://merck.github.io/r2rtf/articles/rtf-row.html#border-type).

In this example, we also demonstrate the possibility to add multiple column headers.

```{r}
head(tbl) %>%
  rtf_colheader(
    colheader = " | Treatment",
    col_rel_width = c(3, 6)
  ) %>%
  rtf_colheader(
    colheader = "Adverse Events | Placebo | Xanomeline High Dose | Xanomeline Low Dose",
    border_top = c("", "single", "single", "single"),
    col_rel_width = c(3, 2, 2, 2)
  ) %>%
  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%
  rtf_encode() %>%
  write_rtf("tlf/intro-ae7.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/intro-ae7.pdf")
```

In the r2rtf [get started](https://merck.github.io/r2rtf/articles/r2rtf.html) page
there are more examples to illustrate how to customize

- title, subtitle
- footnote, data source
- special character
- etc

Those features will be introduced when we first use them in the rest of chapters.

There are other R packages that can achieve similar goals.
Interested reader can refer "Microsoft/LibreOffice Formats" section in 
[CRAN Task View: Reproducible Research](https://cran.r-project.org/web/views/ReproducibleResearch.html)
for an overview.
